---
title: "Group 5"
author: "DB, AS, EF, PS"
date: "2023-07-26"
output: html_document
---

```{r load packages}
library(here)
library(readr)
library(ggplot2)
library(dplyr)
library(McMasterPandemic)
library(rstan)
library(tidyr)
library(lubridate)
```

## Ontario data of reported covid cases between Feb-Mar 2020
```{r load data}
# COVID-19 case data for Ontario dates pre 1 April 2020
source(here::here("team-5/utilities.R"))  #load packages and functions
source(here::here("team-5/models.R")) #load basic model
source(here::here("team-5/pull_data.R")) #load data

#S = 14686092 #this is Quarterly 1 population for 2020, S = 14718133 #for Q2 2020
```

##test the basic SEIR model given to us with Ontario population
```{r set up parameters for "null model"}
#data prep

#prep simulation data
# 
# # initial state, in units of individuals
state = c(
   S = 14865000, #S = susceptible
   E = 0,  #E = exposed,
   I = 1,  #I = infected, first date of infection/known exposure is 2020-02-08
   R = 0 #R = recovered
)

# parameters
params = c(
  beta = 0.7,
  N = sum(state),
  alpha = 0.05,
  gamma = 0.06
)

start_date = "2020-03-01" #first day of exposure is 2020-02-08
end_date = "2020-03-31" #for first


```

## set up seir
```{r}
### seir model set up ###
seir = flexmodel(
  params = params,
  state = state,
  start_date = start_date,
  end_date = end_date,
  do_hazard = TRUE # an option to ensure computations are stable when any of the state variables approach zero (keep this switched on!)
)

### adding rate changes to seir model ###
seir = (
  seir
  %>% add_rate(
    "S", # from compartment S
    "E", # to compartment I
    ~ (1/N) * (beta) * (I) # expression for the flow in terms of parameters and state variables, as a formula
  )
  %>% add_rate("E", "I", ~ (alpha))
  %>% add_rate("I", "R", ~ (gamma))     
)

### stochastic ###
seir_obs_err = (seir
                %>% update_error_dist(
                  S ~ poisson(),
                  E ~ poisson(),
                  I ~ poisson(),
                  R ~ poisson()
                )
)

```

## test our basic SEIR model
```{r test the model}

seir_result = simulation_history(seir)

head(seir_result)


plot_sim(seir, seir_result,
         title = "Deterministic SEIR simulation")


set.seed(15)

seir_obs_err_result = (seir_obs_err
                       # turn on observation error in simluation
                       %>% simulation_history(obs_error = TRUE)
)

plot_sim(seir_obs_err, seir_obs_err_result,
         title = "Stochastic SEIR simulation")

# update model to include incidence
seir_obs_err_inc = (seir_obs_err
                    # add an expression to calculate in the simulation report
                    %>% add_sim_report_expr("incidence", ~ (S_to_E) * (S))
                    # add_error_dist instead of update_ because 
                    # update_ replaces previously attached error distributions
                    # add_ appends
                    %>% add_error_dist( 
                      incidence ~ poisson()
                    )
)

seir_obs_err_inc_result = (seir_obs_err_inc
                           %>% simulation_history(obs_error = TRUE)
)

#plot stochastic
plot_sim(seir_obs_err_inc, seir_obs_err_inc_result,
         title = "Stochastic SEIR simulation with incidence")


# create simulated data

observed = (
  seir_obs_err_inc_result
  %>% select(-matches("to"))
  %>% rename(date = Date)
  %>% pivot_longer(-date, names_to = "var")
  # keep incidence observations
  %>% filter(var == "incidence")
  # lob off first observation 
  # (fitting the initial value is technically difficult and not important to figure out here)
  %>% slice(-1)
)

head(observed)

head(cases)

cases2 <- cases %>% 
  mutate(var = "incidence") %>% 
  rename(value = new_cases) %>% 
  relocate(value, .after = var) %>% filter(date >= "2020-02-09")



# calibrate model to mock data

seir_obs_err_inc_to_calibrate = (seir_obs_err_inc
                                 # attach observed data
                                 %>% update_observed(
                                   cases2
                                 )
                                 # attach priors for parameters we're fitting
                                 # ("optimizing" over)
                                 %>% update_opt_params(
                                   # fitting log beta
                                   log_beta ~ log_normal(
                                     -1, # log mean, so mean beta is exp(-1) = 0.36
                                     0.5 # standard deviation of the log normal prior
                                   )
                                 )
)



model_fit = calibrate_stan(
  model = seir_obs_err_inc, # original model object
  model_to_calibrate = seir_obs_err_inc_to_calibrate, # model object with observed data and priors
  chains = 2 # number of MCMC chains
)

# now look at model fit:

fit = tidy_fit_stan(model_fit)$fit # a simple utility to attach parameter names to stan output
rstan::summary(fit)$summary
rstan::traceplot(fit, ncol = 1)

exp(rstan::summary(fit)$summary["log_beta", "mean"])
params[["beta"]]

# model fit
fit_ensemble_summary = (model_fit
                        %>% ensemble_stan(n_cores = 4) # generate ensemble in parallel
                        %>% summarise_ensemble_stan()
)

head(fit_ensemble_summary)

plot_ensemble(fit_ensemble_summary, cases2)


# forecast:

fcst_ensemble_summary = (model_fit
                         %>% ensemble_stan(
                           days_to_forecast = 30, # new! number of days to forecast
                           n_cores = 4
                         )
                         %>% summarise_ensemble_stan() # can specify a different quantile vector here: see documentation
)

plot_ensemble(fcst_ensemble_summary, cases2)

View(fcst_ensemble_summary)

## paper pre Apr 2020: chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/https://www.medrxiv.org/content/10.1101/2020.03.24.20042705v1.full.pdf


``` 

## looks like overfitting is occurring, need to improve model; check parameters
## compare higher and lower betas

```{r sensitivity analysis}


```

## compare different betas to see how this affects our distribution
```{r fit gamma}

```


```{r fit alpha}
```

## consider our objective: mask effect; fit posteriors 

```{fit posteriors}

```