---
title: "Fitting COVID-19 case data using an SEIR model in STAN"
format: 
  html:
    fig-width: 8
    fig-height: 6
editor: 
  markdown: 
    wrap: 72
---

```{r knit options}
#| include: false
options(future.rng.onMisuse = "ignore")
options(mc.cores = parallel::detectCores())
```

Set-up Stan and related libraries, as well as libraries for data and
data visualization.

```{r setup, message=FALSE, warning=FALSE}
#| message: false
#| warning: false
# Bayesian computation
library(rstan)
library(tidybayes)

# data science packages
library(dplyr)
library(magrittr)
library(ggplot2)
library(readr)
library(here)
ymd <- lubridate::ymd
```

# Data

First, we will read in the British Columbia, Canada COVID-19
reported-case data. 

## Load the BC COVID-19 data

```{r load bc data}
# load data grouped by location (HA), and age group
#covid_data <- read_csv("/Users/johnmensah/Documents/Ecological\ Forecasting_Workshop/Team\ 4/CaseStudy_Covid19/covidseir/data/fitting_data.csv")

covid_data <- read_csv(here("Documents","data", "fitting_data.csv")) 
covid_data_val <- read_csv(here("Documents","data", "validation_data.csv"))

plot_covid_both <- bind_rows(covid_data, covid_data_val) %>% 
  group_by(Reported_Date, HA) %>%
  tally() %>%
  ggplot(aes(x = Reported_Date, y = n, color = HA)) +
  geom_line() +
  theme_classic()



# Load Total BC cases for whole province for first 60 recorded time-points
#bc_data <- read_csv("/Users/johnmensah/Documents/Ecological\ Forecasting_Workshop/Team\ 4/CaseStudy_Covid19/covidseir/data/bc_fitting_data.csv")

bc_data  <- read_csv(here("Documents","data", "bc_fitting_data.csv")) %>% 
  mutate(data= "fit")
bc_data_val  <- read_csv(here("Documents","data", "bc_validation_data.csv"))%>% 
  mutate(data= "validation")

bc_data_both <- bind_rows(bc_data, bc_data_val)

plot.bc.data <- bc_data_both%>% 
  ggplot(aes(x = Reported_Date, y = n, color= data)) +
  geom_vline(xintercept=ymd("2020-03-28"))+
  geom_line() +
  theme_classic() +
  ggtitle("Reported cases BC (first 60 recorded time-points of pandemic)")


plot.bc.data 

ggsave("plot.bc.data.jpeg", plot.bc.data  )


```


The task will be to use the first 30 recorded time-points of data to forecast 
the next 30 days of cases.
In order to do this we first need to define over what days to forecast for. As
stan code can't handle dates in as clean a way that can be done in R, we will 
instead generate the days to forecast as the number of days since the start of
the data

We then store the time points and cases data to load into the stan model

```{r store fitting forecast data}
fitting_time_points <- bc_data$time_points
fitting_cases <- bc_data$n
forecast_time_points <- seq(max(fitting_time_points) + 1,
                            max(fitting_time_points) + 31)

forecast_dates <- seq(max(bc_data$Reported_Date) + 1,
                      max(bc_data$Reported_Date) + 31, by = "1 day")
```

# Setting up the model

We will start with a simple SEIR model with the following structure

$$
\begin{aligned}
\frac{{dS}}{{dt}} &= -\beta \cdot S \cdot \frac{{I}}{{N}} \\
\frac{{dE}}{{dt}} &= \beta \cdot S \cdot \frac{{I}}{{N}} - \sigma \cdot E \\
\frac{{dI}}{{dt}} &= \sigma \cdot E - \gamma \cdot I \\
\frac{{dR}}{{dt}} &= \gamma \cdot I
\end{aligned}
$$

In these equations, the variables represent the following:

- $S$ is the number of susceptible individuals.
- $E$ is the number of exposed individuals (infected but not yet infectious).
- $I$ is the number of infectious individuals.
- $R$ is the number of recovered (or removed) individuals.
- $N$ is the total population size.
- $\beta$ is the infection rate (rate of transmission from infectious to susceptible individuals).
- $\sigma$ is the incubation rate (rate of transition from the exposed to infectious state).
- $\gamma$ is the recovery rate (rate at which infectious individuals recover or are removed).

The $R_0$ can be directly derived from the above model by considering the 
expected number of secondary infections following a primary infection in a large
population. Since the expected infectious time is $1/\gamma$ and the rate of 
infections in one unit time is $\beta$, then $R_0 = \beta/\gamma$. [Note that
this assumes a large population since in order to derive this result we assume
that $N - 1 \approx N$].

We start by assuming the incubation rate and recovery rate are fixed and fitting
the $R_0$ and the initial number of infected individuals $I_0$. The number of
infected are assumed to have a log-normal distribution,

$$\log(I_0) \sim \mathcal{N}(\mu_{I0},\sigma_{I0}),$$ 
In addition, the $R_0$ is assumed to be normally distributed on the positive
real half,

$$R_0 \sim \mathcal{N}(\mu_{R0},\sigma_{R0}).$$

Finally for the observation process it is assumed that the number of cases
observed are Poisson distributed with a rate that is a fixed proportion of the 
total number of infected individuals at a given time-point. The sampled fraction
of cases ($s$) is assumed to be bounded between 0 and 1 with a mean 0.2 and standard
deviation 0.01,
$$s \sim \mathcal{N}(0.2,0.01)[0,1].$$

Finally the likelihood can be written as,

$$C_t \sim \text{Poisson}(sI_t).$$

# Loading the stan model

This vignette uses a fairly bare bones SEIR model implementation with a
Poisson observation process. The following chunk
prints the lines contained in the `stan` file

```{r read stan model}
#code <- readLines("/Users/johnmensah/Documents/Ecological\ Forecasting_Workshop/Team\ 4/CaseStudy_Covid19/covidseir/stan_models/basic_seir_model.stan")

code <- readLines(here("Documents","stan_models", "basic_seir_model.stan"))

cat(code, sep = "\n")

```

We run the following to compile the model

```{r compile stan model}
#model <- stan_model("/Users/johnmensah/Documents/Ecological\ Forecasting_Workshop/Team\ 4/CaseStudy_Covid19/covidseir/stan_models/basic_seir_model.stan")

model <- stan_model(here("Documents","stan_models", "basic_seir_model.stan"))
```

Note what is provided in the "data" block of the Stan code. That is what
is expected as input at inference time. This can be provided as a `list`
object.

```{r load data into stan format}
stan_data <- list(
  T = length(fitting_time_points), # number of data points
  y = fitting_cases, # observed infection cases
  ts = fitting_time_points, # Time points
  forecast_T = length(forecast_time_points), # number of forecast points
  forecast_ts = forecast_time_points, # Forecast time points
  R0_prior = c(log(2.5), 0.2), # log Mean and std for R0
  i0_prior = c(log(8), 1.0), # log Mean and std for i0
  gamma = 1 / 7, # recovery rate
  sigma = 1 / 5, # incubation rate
  pop_size = 5e6, # population of BC
  t0 = -1
)
```

# Run the inference

```{r}
fit <- sampling(model,
  data = stan_data,
  iter = 1000,
  seed = 42, # fix seed to recreate results
  chains = 2,
  control = list(adapt_delta = 0.9)
)


#########Extended Model##############
#fit_extened <- sampling(model_extended,
 # data = stan_data,
 # iter = 500,
 # seed = 42, # fix seed to recreate results
 # chains = 1,
 # control = list(adapt_delta = 0.9)
#)
```

Examine the summary of the posterior

```{r}
summary(fit, c("R0", "i0", "sample_frac"))
```

# Examining the model output

Compare the output of the different states in the model,

```{r}
state_dictionary <- tribble(
  ~i, ~state,
  1, "S",
  2, "E",
  3, "I",
  4, "R"
)

time_dictionary <- bc_data %>%
  select(Reported_Date) %>%
  mutate(t = 1:n())

########## Infected Individuals###########
fit %>%
  spread_draws(y_hat[t, i]) %>%
  left_join(state_dictionary, by = "i") %>%
  left_join(time_dictionary, by = "t") %>%
  filter(state == "I") %>%
  ggplot(aes(x = Reported_Date)) +
  stat_lineribbon(aes(y = y_hat), .width = c(.99, .95, .8, .5)) +
  scale_fill_brewer() + ylab("Infected Individuals") + theme_bw()


############### Exposed individuals##################
fit %>%
  spread_draws(y_hat[t, i]) %>%
  left_join(state_dictionary, by = "i") %>%
  left_join(time_dictionary, by = "t") %>%
  filter(state == "E") %>%
  ggplot(aes(x = Reported_Date)) +
  stat_lineribbon(aes(y = y_hat), .width = c(.99, .95, .8, .5)) +
  scale_fill_brewer() + ylab("Exposed Indiviuals") + theme_bw()


############### Recovered Individuals##################
fit %>%
  spread_draws(y_hat[t, i]) %>%
  left_join(state_dictionary, by = "i") %>%
  left_join(time_dictionary, by = "t") %>%
  filter(state == "R") %>%
  ggplot(aes(x = Reported_Date)) +
  stat_lineribbon(aes(y = y_hat), .width = c(.99, .95, .8, .5)) +
  scale_fill_brewer() + ylab("Recovered Individuals") + theme_bw()

############### Susceptible Individuals##################
fit %>%
  spread_draws(y_hat[t, i]) %>%
  left_join(state_dictionary, by = "i") %>%
  left_join(time_dictionary, by = "t") %>%
  filter(state == "S") %>%
  ggplot(aes(x = Reported_Date)) +
  stat_lineribbon(aes(y = y_hat), .width = c(.99, .95, .8, .5)) +
  scale_fill_brewer() + ylab("Susceptible Individuals") + theme_bw()
```

Compare posterior predictive distribution of cases to actual cases,

```{r}
time_dictionary <- bc_data_both %>%
  select(Reported_Date) %>%
  mutate(t = 1:n())

plot.original.model.fit.validation <- fit %>%
  spread_draws(cases[t]) %>%
  left_join(time_dictionary, by = "t") %>%
  ggplot(aes(x = Reported_Date)) +
  stat_lineribbon(aes(y = cases), .width = c(.99, .95, .8, .5), color = "#08519C") +
  scale_fill_brewer() +
  geom_point(data = bc_data_both, aes(x = Reported_Date, y = n, color= data))+
  theme_bw()

ggsave("plot.original.model.fit.validation.jpeg", plot.original.model.fit.validation)


```

```{r}
time_dictionary <- bc_data %>%
  select(Reported_Date) %>%
  mutate(t = 1:n())

plot.original.model.fit <- fit %>%
  spread_draws(cases[t]) %>%
  left_join(time_dictionary, by = "t") %>%
  ggplot(aes(x = Reported_Date)) +
  stat_lineribbon(aes(y = cases), .width = c(.99, .95, .8, .5), color = "#08519C") +
  scale_fill_brewer() +
  geom_point(data = bc_data, aes(x = Reported_Date, y = n))

ggsave("plot.original.model.fit.jpeg", plot.original.model.fit)


```

Compare the forecasted predictive distribution of cases to actual cases,

```{r}
time_dictionary_fitting <- bc_data %>%
  select(Reported_Date) %>%
  mutate(
    t = 1:n(),
    .variable = "cases"
  )
time_dictionary_forecasting <- tibble(Reported_Date = forecast_dates) %>%
  mutate(
    t = 1:n(),
    .variable = "forecasted_cases"
  )

time_dictionary <- time_dictionary_fitting %>%
  bind_rows(time_dictionary_forecasting)

plot.original.model.fit.validation <-fit %>%
  tidybayes::gather_draws(cases[t], forecasted_cases[t]) %>%
  left_join(time_dictionary, by = c("t", ".variable")) %>%
  ggplot(aes(x = Reported_Date)) +
  stat_lineribbon(aes(y = .value, group = .variable),
    color = "#08519C",
    .width = c(.8, .5)
  ) +
  ylim(0,200)+
  scale_fill_brewer() +
  geom_point(data = bc_data_both, aes(x = Reported_Date, y = n, color= data))+
  theme_bw()


ggsave("plot.original.model.fit.validation.jpeg", plot.original.model.fit.validation)


```

The base model indicates a period of unbounded growth. In addition,
The posterior predictive distribution of cases does not seem to match the
pattern of observed cases. Bayesian p-values and residuals can be used to 
investigate these differences. What are some of the underlying causes of this? 

```{r}

data.com <- fit %>%
  tidybayes::gather_draws(cases[t]) %>%
  left_join(time_dictionary, by = c("t", ".variable")) %>%
  left_join(bc_data, by = "Reported_Date") %>%
  mutate(residuals = n - .value)

plot.residuals.original.model <- data.com  %>%
  filter(.draw %in% c(2,4,5,6,7,8,9,10)) %>% 
  ggplot(aes(x = t, y=residuals, color = as.factor(.draw))) + 
  geom_point() + 
  theme_bw() +
  geom_hline(yintercept = 0, linetype="dashed")


ggsave("plot.residuals.original.model.jpeg", plot.residuals.original.model)

```

Look at diagnostics in shinystan

```{r}
y_data <- bc_data$n
shinystan::launch_shinystan(fit)
```

Do mean plot of residuals

```{r}
data.com  %>%
  group_by(t) %>% 
  summarise(mean_residuals = mean(residuals)) %>% 
  ggplot(aes(x = t, y=mean_residuals)) + 
  geom_point() + 
  theme_bw() +
  geom_hline(yintercept = 0, linetype="dashed")

```

# Decision maker request

Policy makers are considering a number of potential public health measures to 
reduce the transmission of COVID-19 and the number of cases. The main measure 
currently being considered is requiring masking in indoor public settings

What would be the projected number of cases over the next 30 days under this 
scenario? If the intervention is not introduced, what would be the projected 
number of cases over the next 30 days?

In addition to the outlined questions, further directions for the case
study, which may enhance the model forecast are given below.

# Model extensions

## Reported cases

The number of reported cases is not necessarily a fixed relationship
with the number of infections. Factors such as the severity of cases,
the age distribution of infections, and testing policies among others
will play a role in what infections are detected. Expand the model to
consider the observation process of generated cases from infections. Could this
process be over-dispersed? In addition consider whether this varies over time by
different factors and whether observations are over-dispersed.

## Adding covariates

The rate of transmission depends on many factors such as the rate of
contact within the population, including the setting and type of
contact. The observation of cases from infections also depends on many
factors including weekday/weekend, which may impact when testing
locations are open and people's ability to seek testing. Consider how to
extend the model to include covariates for the transmission rate and
testing rate. What implications for public health policy can be made by
these inferences? This may involve the incorporation of external data, such as
policy stringency measures or measures of mobility. See the 
[Our World in Data page](https://ourworldindata.org/policy-responses-covid) for 
some potential resources.

## Contact rate breakpoints

Changes in public policy and population behavior impacts the
transmission rate. Further policies such as self-isolation following
symptom onset or testing positive can additionally impact the expected
time that an individual is infectious. Expand the model to consider
varying behavior and policy. Can changes in behavior be inferred from
these case data?

## Geographic differences

Population structure is important when considering where an outbreak
emerges and how it develops. Extend the model to incorporate different
health regions. What would be the impact of introducing public health
measures that differentially impacts different populations?

## Population age structure

Age structure is often considered through the
who-aquires-infection-from-whom matrix (WAIFM). Extend the model to
incorporate age structure and consider how measures such as those which
impact school-aged children would impact the total observed numbers of
cases.
