---
title: "NEON forecast challenge - chlorophyll"
author: Freya Olsson, Quinn Thomas
output:
  md_document: 
    variant: markdown_github
    number_sections: true
    toc: true
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# This R markdown document
This is a vignette on producing surface lake chlorophyll forecasts using NEON data. Data are used in the NEON Forecasting Challenge and these materials have been modified based on an original workshop materials focused on submitting to the Challenge. To complete the workshop via this markdown document the following packages will need to be installed:

* `remotes`
* `rjags`
* `tidybayes`
* `tidyverse`
* `lubridate`
* `neon4cast` (from github)

For the rjags code to work you first you need to install JAGS code from: [https://mcmc-jags.sourceforge.io](https://mcmc-jags.sourceforge.io)

The following code chunk should be run to install packages. If you are using the eco4cast/neon4cast Docker container you will not need to do this install step. 

```{r eval = F}
install.packages('remotes')
install.packages('rjags')
install.packages('tidybayes')
install.packages('tidyverse') # collection of R packages for data manipulation, analysis, and visualisation
install.packages('lubridate') # working with dates and times
remotes::install_github('eco4cast/neon4cast') # package from NEON4cast challenge organisers to assist with forecast building and submission
```

Additionally, R version 4.2 is required to run the neon4cast package. It's also worth checking your Rtools is up to date and compatible with R 4.2, see (https://cran.r-project.org/bin/windows/Rtools/rtools42/rtools.html). 


```{r}
version$version.string
library(tidybayes)
library(tidyverse)
library(lubridate)
library(rjags)

```

If you do not wish to run the code yourself you can follow along via the rendered markdown which can be viewed from the [Github repository](https://github.com/OlssonF/NEON-forecast-challenge-workshop). 

# The case study

In this water quality forecasting example we will produce 30 day ahead forecasts of lake surface chlorophyll concentration. Chlorophyll concentration is a proxy used in water quality monitoring for algal biomass. Large chlorophyll values can be an indication of algal blooms. Algal blooms can be a problem for managers as they are linked to low oxygen concentrations, fish kills and toxin production, all important water quality parameters. The concentration of chlorophyll in natural waters is driven by a complex interaction of temperatures, nutrients, and hydrodynamic conditions that is not well quantified and can be quite stochastic in nature - a great challenge for forecasters!

# The forecasting workflow
## Read in the data

We start forecasting by first looking at the historic data - called the 'targets'. These data are available near real-time, with the latency of approximately 24-48 hrs but for this introduction we will look at a subset of the historic data.

```{r eval=TRUE, echo = TRUE, error=FALSE, warning=FALSE, message=FALSE}
#read in the targets data
targets <- read_csv('Data/targets-neon-chla.csv')
target_sites <- targets |> 
  distinct(site_id) |> 
  pull()

```

Information on the NEON sites can be found in the `NEON_Field_Site_Metadata_20220412.csv` file on the eco4cast GitHub. It can be filtered to only include the sites we will be looking at. This table has information about the field sites, including location, ecoregion, information about the watershed (e.g. elevation, mean annual precipitation and temperature), and lake depth. 
 
```{r eval=TRUE, echo = TRUE, error=FALSE, warning=FALSE, message=FALSE}
# read in the sites data
aquatic_sites <- read_csv("https://raw.githubusercontent.com/eco4cast/neon4cast-targets/main/NEON_Field_Site_Metadata_20220412.csv") |>
  dplyr::filter(field_site_id %in% target_sites)
```

Let's take a look at the targets data!
```{r eval = T, echo = TRUE}
glimpse(targets)

```

The columns of the targets file show the time step (daily for this water quality data), the 4 character site code (`site_id`), the variable being measured, and the mean daily observation. Here we are looking only at two lake sites (`BARC` and `SUGG`) chlorophyll concentration (ug/L) (`chla`) observations.

## Visualise the data
```{r eval = TRUE, echo = TRUE, warning=FALSE, fig.dim=c(10,10), fig.cap='Figure: Chlorophyll targets data at two lake sites provided by EFI for the NEON forecasting challgenge.'}

targets %>%
  ggplot(., aes(x = datetime, y = observation)) +
  geom_point() +   
  theme_bw() + theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  facet_wrap(~site_id, scales = 'free_y')+
  labs(title = 'chla')

```


# Introducing co-variates

One important step to overcome when thinking about generating forecasts is to include co-variates in the model. A chlorophyll forecast, for example, may be benefit from information about past and future weather. Below is code snippets for downloading past and future NOAA weather forecasts for all of the NEON sites. The 3 types of data are as follows:

* stage_1: raw forecasts - 31 member ensemble forecasts at 3 hr intervals for the first 10 days, and 6 hr intervals for up to 35 days at the NEON sites.
* stage_2: a processed version of Stage 1 in which fluxes are standardized to per second rates, fluxes and states are interpolated to 1 hour intervals and variables are renamed to match conventions. We recommend this for obtaining future weather. Future weather forecasts include a 30-member ensemble of equally likely future weather conditions.
* stage_3: can be viewed as the "historical" weather and is combination of day 1 weather forecasts (i.e., when the forecasts are most accurate). 

This code create a connection to the dataset hosted on the eco4cast server. To download the data you have to tell the function to `collect()` it. These data set can be subsetted and filtered using `dplyr` functions prior to download to limit the memory usage.

You can read more about the NOAA forecasts available for the NEON sites [here:](https://projects.ecoforecast.org/neon4cast-docs/Shared-Forecast-Drivers.html)

## Download co-variates
### Download historic data

We will generate a chlorophyll forecast using `air_temperature` as a co-variate. 
Note: This code chunk can take a little to execute as it accesses the NOAA data depending on your internet connection.

```{r, message=FALSE}
# past stacked weather
bucket <- "neon4cast-drivers/noaa/gefs-v12/stage3/parquet/"
s3_past <- arrow::s3_bucket(bucket, endpoint_override = "data.ecoforecast.org", anonymous = TRUE)

variables <- c("air_temperature")
#Other variable names can be found at https://projects.ecoforecast.org/neon4cast-docs/Shared-Forecast-Drivers.html#stage-3

noaa_past <- arrow::open_dataset(s3_past) |> 
  dplyr::filter(site_id %in% target_sites,
                datetime >= ymd('2017-01-01'),
                variable %in% variables) |> 
  collect()

noaa_past[1:10,]
```

This is a stacked ensemble forecast of the one day ahead forecasts. To get an estimate of the historic conditions we can take a mean of these ensembles. We will also need to convert the temperatures to Celsius from Kelvin.

```{r}
# aggregate the past to mean values
noaa_past_mean <- noaa_past |> 
  mutate(datetime = as_date(datetime)) |> 
  group_by(datetime, site_id, variable) |> 
  summarize(prediction = mean(prediction, na.rm = TRUE), .groups = "drop") |> 
  pivot_wider(names_from = variable, values_from = prediction) |> 
  # convert air temp to C
  mutate(air_temperature = air_temperature - 273.15) |> 
  filter(datetime <= as_date(max(targets$datetime)))

```

We can then look at the future weather forecasts in the same way but using the `noaa_stage2()`. The forecast becomes available from NOAA at 5am UTC the following day, so if we were producing realtime forecasts we would need take the NOAA forecast from yesterday to make the water quality forecasts. But for this example we will produce forecasts from the end of the targets for the next 35 days. The stage 2 NOAA forecast is all an ensemble of 31 simulations which we can use to produce an estimate of driver uncertainty in the chlorophyll forecast. 

### Download future weather forecasts

```{r, message=FALSE}
# Note: New forecast only available at 5am UTC the next day - so if forecasting in realtime look for yesterday's forecast

# generate forecast(s) for June 2023
forecast_date <- as_date(max(targets$datetime))

bucket <- paste0("neon4cast-drivers/noaa/gefs-v12/stage2/parquet/0/", forecast_date) 
s3_future <- arrow::s3_bucket(bucket = bucket, endpoint_override = 'data.ecoforecast.org', anonymous = TRUE)

variables <- c("air_temperature")

noaa_future <- arrow::open_dataset(s3_future) |> 
  dplyr::filter(datetime >= forecast_date,
                site_id %in% target_sites,
                variable %in% variables) |> 
  collect()

```

The forecasts are hourly and we are interested in using daily mean air temperature for lake chlorophyll forecast generation.

```{r warning=F}
noaa_future_daily <- noaa_future |> 
  mutate(datetime = as_date(datetime)) |> 
  # mean daily forecasts at each site per ensemble
  group_by(datetime, site_id, parameter, variable) |> 
  summarize(prediction = mean(prediction)) |>
  pivot_wider(names_from = variable, values_from = prediction) |>
  # convert to Celsius
  mutate(air_temperature = air_temperature - 273.15) |> 
  select(datetime, site_id, air_temperature, parameter)

noaa_future_daily[1:10,]
```
Now we have a timeseries of historic data and a 30 member ensemble forecast of future air temperatures.

```{r echo = T, fig.cap = c('Figure: historic and future NOAA air temeprature forecasts at lake sites', 'Figure: last two months of historic air temperature forecasts and 35 day ahead forecast')}
ggplot(noaa_future_daily, aes(x=datetime, y=air_temperature)) +
  geom_line(aes(group = parameter), alpha = 0.4)+
  geom_line(data = noaa_past_mean, colour = 'darkblue') +
  facet_wrap(~site_id, scales = 'free')

ggplot(noaa_future_daily, aes(x=datetime, y=air_temperature)) +
  geom_line(aes(group = parameter), alpha = 0.4)+
  geom_line(data = noaa_past_mean, colour = 'darkblue') +
  coord_cartesian(xlim = c(forecast_date - days(60),
                           forecast_date + days(35)))+
  facet_wrap(~site_id, scales = 'free')
```

# Model fitting

We will fit a simple dynamic JAGS model with historic air temperature and the chlorophyll targets data. Using this model we can then use our future estimates of air temperature (all 30 ensembles) to forecast chlorophyll at each site. The model will estimate both process and observational uncertainty. In addition, the ensemble weather forecast will also give an estimate of driver data uncertainty. 

We will start by joining the historic weather data with the targets to aid in fitting the model.

```{r}
targets_lm <- targets |> 
  pivot_wider(names_from = 'variable', values_from = 'observation') |> 
  left_join(noaa_past_mean, 
            by = c("datetime","site_id"))

targets_lm[1050:1060,]
```


```{r}
example_site <- 'SUGG'

site_target <- targets_lm |> 
  filter(site_id == example_site)

#Find when the data for the site starts and filter to only 
#more recent datetimes with no NAs
first_no_na <- site_target |> 
  filter(!is.na(air_temperature) & !is.na(chla)) |> 
  summarise(min = min(datetime)) |> 
  pull(min)

site_target <- site_target |> 
  filter(datetime >= first_no_na)
```

We fit the model as a state-space Bayesian model.  The following is BUGS code  that specifies the model. Explaining the Bayesian model is beyond the scope of this tutorial. 

```{r }
jags_code <- "
model{

  # Priors
  beta1 ~ dnorm(0, 1/10000)
  beta2 ~ dnorm(0, 1/10000)
  sd_process ~ dunif(0.00001, 100)
  
  #Convert Standard Deviation to precision
  tau_obs <- 1 / pow(sd_obs, 2)
  tau_process <- 1 / pow(sd_process, 2)

  #Initial conditions
  chla_latent[1] <- chla_init
  y[1] ~ dnorm(chla_latent[1], tau_process)  

  #Loop through data points
  for(i in 2:n){
      # Process model
      chla_pred[i] <- beta1 * chla_latent[i-1] + beta2 * air_temp[i]
      chla_latent[i] ~ dnorm(chla_pred[i], tau_process)

      # Data model
      y[i]  ~ dnorm(chla_latent[i], tau_obs)
  }
}
"
```

Set up the fixed inputs for the JAGS model specification.  This includes the covariates, observations, loop indexes, initial conditions, and any fixed parameter values.  Here we fix the standard deviation of the observations, since it is challenging to estimate without multiple measurements at the same datetime.

```{r}
data <- list(air_temp = site_target$air_temperature,
             y = site_target$chla,
             n = length(site_target$air_temperature),
             sd_obs = 0.1,
             chla_init = site_target$chla[1])
```

Initialize random variables in JAGS model for each of the 3 chains.

```{r}
nchain <- 3
inits <- list()
for(i in 1:nchain){
  inits[[i]] <- list(beta1 = rnorm(1, 0.34, 0.05), 
                     beta2 = rnorm(1, 0.11, 0.05),
                     sd_process = runif(1, 0.05, 0.15 ))
}
```

Build JAGS model object and fit JAGS code

```{r}
j.model <- jags.model(file = textConnection(jags_code),
                      data = data,
                      inits = inits,
                      n.chains = 3)

#Run MCMC
jags.out   <- coda.samples(model = j.model,
                           variable.names = c("beta1", 
                                              "beta2",
                                              "chla_latent",
                                              "sd_process"),
                           n.iter = 5000)
```

Convert MCMC object to tidy data and remove iterations in burn-in.  We only keep the latent state on last day for use as initial conditions in the forecast.

```{r warning=F}

burn_in <- 1000
chain <- jags.out %>%
  tidybayes::spread_draws(beta1, beta2, sd_process, chla_latent[day]) |> 
  filter(.iteration > burn_in) |> 
  ungroup()

max_day <- max(chain$day)
chain <- chain |> 
  filter(day == max_day) |> 
  select(-day)
```

Visualize the chain to inspect for convergence.

```{r}
chain |> 
  pivot_longer(beta1:chla_latent, names_to = "variable" , values_to = "value") |> 
  ggplot(aes(x = .iteration, y = value, color = factor(.chain))) +
  geom_line() +
  facet_wrap(~variable, scale = "free")
```

## Forecast with posteriors

Here we generate a forecast using 500 random draws from the posterior distributions.  The following code loops over the samples and then forecasts each day in the future.  It also samples from the ensembles in the NOAA weather forecast. 

The forecasted values is the latent state forecast with observation uncertainty added.

```{r}
num_samples <- 500

noaa_future_site <- noaa_future_daily |> 
  filter(site_id == example_site)

n_days <- length(unique(noaa_future_site$datetime))
chla_latent <- matrix(NA, num_samples, n_days)   
y <- matrix(NA, num_samples, n_days) #y is the forecasted observation
forecast_df <- NULL



#loop over posterior samples
for(i in 1:num_samples){
  sample_index <- sample(x = 1:nrow(chain), size = 1, replace = TRUE)
  noaa_ensemble <- sample(x = 1:30, size = 1)
  noaa_future_site_ens <- noaa_future_site |> filter(parameter == noaa_ensemble)
  #Initialize with a sample from the most recent latent state posterior distribution
  chla_latent[i,1] <- chain$chla_latent[sample_index]
  y[i, 1] <- rnorm(1, chla_latent[i,1], sd = data$sd_obs)
  
  #loop over forecast days
  for(j in 2:n_days){
    pred <- chla_latent[i,j-1] * chain$beta1[sample_index] +
      noaa_future_site_ens$air_temperature[j] * chain$beta2[sample_index]
    
    chla_latent[i,j] <- rnorm(1, pred, chain$sd_process[sample_index])
    y[i,j] <- rnorm(1, chla_latent[i,j], sd = data$sd_obs)
    
  }
  
  #Build formated forecast
  df <- tibble(datetime = noaa_future_site_ens$datetime,
               site_id = example_site,
               variable = "chla",
               parameter = i,
               prediction = y[i, ])
  
  forecast_df <- bind_rows(forecast_df, df)
}
```

We now have 500 possible trajectories of chlorophyll-a  at each site and each day. On this plot each line represents one of the possible trajectories and the range of forecasted trajectories is a simple quantification of the uncertainty in our forecast.

Looking back at the forecasts we produced:

```{r, echo = T, warning = F}
forecast_df %>% 
  ggplot(.,aes(x=datetime, y=prediction, group = parameter)) + 
  geom_point(data = subset(targets, site_id == 'SUGG'),
             aes(x=datetime, y=observation, group = 'obs'), colour = 'black') +
  geom_line(alpha = 0.3, aes(colour = 'ensemble member (parameter)')) + 
  facet_wrap(~site_id, scales = 'free_y') +
  scale_x_date(expand = c(0,0), date_labels = "%d %b") +
  labs(y = 'value') +
  geom_vline(aes(linetype = 'reference_datetime', 
                 xintercept = as_date(max(targets$datetime))), 
             colour = 'blue', size = 1.5) +
  labs(title = 'site_id', subtitle = 'variable = temperature', caption = 'prediction') + 
  annotate("text", x = min(forecast_df$datetime) - days(10), y = 40, label = "past")  +
  annotate("text", x = min(forecast_df$datetime) + days(12), y = 40, label = "future")  +
  theme_bw() +
  coord_cartesian(xlim = c(min(forecast_df$datetime) - 30, Sys.Date())) +
  scale_linetype_manual(values = 'dashed', name = '')+
  scale_colour_manual(name = '', values = 'red')
```

## Convert to EFI standard 

Forecast standards are useful for maintaining consistent formats across forecast generations. For an ensemble forecast the standards specifthe y following columns:

* `datetime`: forecast timestamp for each time step
* `reference_datetime`: The start of the forecast; this should be 0 times steps in the future. This should only be one value of reference_datetime in the file
* `site_id`: NEON code for site
* `family`: name of probability distribution that is described by the parameter values in the parameter column; only `normal` or `ensemble` are currently allowed.
* `parameter`: integer value for forecast replicate (from the `.rep` in fable output);
* `variable`: standardized variable name from the theme 
* `prediction`: forecasted value (from the `.sim` column in fable output)
* `model_id`: model name (no spaces)

We need to make sure the dataframe is in the correct format.  This is an ensemble forecast (specified in the `family` column). 

```{r}
# Remember to change the model_id when you make changes to the model structure!
my_model_id <- 'your_model_id'

forecast_df_efi <- forecast_df %>%
  mutate(model_id = my_model_id,
         reference_datetime = as_date(min(datetime)) - days(1),
         family = 'ensemble',
         parameter = as.character(parameter)) %>%
  select(model_id, datetime, reference_datetime, site_id, family, parameter, variable, prediction)
```

For the NEON forecast Challenge (including the aquatics theme) the following standards are needed for a forecast submission. 


# Introduction to NEON forecast challenge

These data collected by NEON and form part of the EFI-NEON Forecasting Challenge. The Challenge asks the scientific community to produce ecological forecasts of future conditions at NEON sites by leveraging NEON's open data products. The Challenge is split into five themes that span aquatic and terrestrial systems, and population, community, and ecosystem processes across a broad range of ecoregions. We are excited to use this Challenge to learn more about the predictability of ecological processes by forecasting NEON data before it is collected.  
  
Which modeling frameworks, mechanistic processes, and statistical approaches best capture community, population, and ecosystem dynamics? These questions are answerable by a community generating a diverse array of forecasts. The Challenge is open to any individual or team from anywhere around the world that wants to submit forecasts. Sign up [here.](https://projects.ecoforecast.org/neon4cast-docs/Participation.html). 

## Aquatics challenge

What: Freshwater surface water temperature, oxygen, and chlorophyll-a.

Where: 7 lakes and 27 river/stream NEON sites.

When: Daily forecasts for at least 30-days in the future. New forecast submissions, that use new data to update the forecast, are accepted daily. The only requirement is that submissions are predictions of the future at the time the forecast is submitted. To submit a forecast to the Challenge these need to be __realtime__ (actual forecasts of the future)!
Real-time data are available for submissions to the Challenge. See [documentation](https://projects.ecoforecast.org/neon4cast-docs/Aquatics.html).

Today we looked at just two of the lake sites and forecasted chlorophyll concentration of a historic time period. For the challenge, you can chose to submit to any/all of the sites! You can also chose to submit any of the three focal variables (temperature, oxygen, and chlorophyll). But should be acutal forecasts of the future! Find more information about the aquatics challenge [here](https://projects.ecoforecast.org/neon4cast-docs/Aquatics.html).

## Submission requirements

For the Challange, forecasts must include quantified uncertainty. The file can represent uncertainty using an ensemble forecast (multiple realizations of future conditions) or a distribution forecast (with mean and standard deviation), specified in the family and parameter columns of the forecast file. 

For an ensemble forecast, the `family` column uses the word `ensemble` to designate that it is a ensemble forecast and the parameter column is the ensemble member number (1, 2, 3 …).  For a distribution forecast, the `family` column uses the word `normal` to designate a normal distribution and the parameter column must have the words mu and sigma for each forecasted variable, site_id, and datetime. For forecasts that don't have a normal distribution we recommend using the ensemble format and sampling from your non-normal distribution to generate a set of ensemble members that represents your distribution. I will go through examples of both `ensemble` and `normal` forecasts as examples. 

The full list of required columns and format can be found in the [Challenge documentation](https://projects.ecoforecast.org/neon4cast-docs/Submission-Instructions.html).

## Optional: Submit forecast

The final step to submit a formatted forecast to the Challenge is below. The forecast organizers have created tools to help aid in the submission process. These tools can be downloaded from Github using `remotes::install_github(eco4cast/neon4cast)`.
These include functions for submitting, scoring and reading forecasts:

* `submit()` - submit the forecast file to the neon4cast server where it will be scored
* `forecast_output_validator()` - will check the file is in the correct format to be submitted
* `check_submission()` - check that your submission has been uploaded to the server

The file name needs to be in the format theme-reference_datetime-model_id

```{r eval = T}
# Start by writing the forecast to file
theme <- 'aquatics'
date <- forecast_df_efi$reference_datetime[1]
forecast_name_1 <- paste0(forecast_df_efi$model_id[1], ".csv")
forecast_file_1 <- paste(theme, date, forecast_name_1, sep = '-')
forecast_file_1

if (!dir.exists('Forecasts')) {
  dir.create('Forecasts')
}
write_csv(forecast_df_efi, file.path('Forecasts', forecast_file_1))

neon4cast::forecast_output_validator(file.path('Forecasts', forecast_file_1))

```

```{r eval = FALSE}
# can uses the neon4cast::forecast_output_validator() to check the forecast is in the right format
neon4cast::submit(forecast_file = file.path('Forecasts', forecast_file_1),
                  ask = FALSE) # if ask = T (default), it will produce a pop-up box asking if you want to submit
```


## Possible modifications to the simple model: 

* Include additional NOAA co-variates in the linear model (remember to 'collect' and subset the right data from NOAA)
* Specify a non-linear relationship
* Include a lag in the predictors

Remember to change the `model_id` so you can differentiate different forecasts!

## Register your participation
It's really important that once you start submitting forecasts to the Challenge that you register your participation. We ask that you complete this [form](https://nd.qualtrics.com/jfe/form/SV_9MJ29y2xNrBOjqZ) which asks you some simple questions about your forecast and team. This is crucial for a couple of reasons:

1. We can keep track different forecast submissions during the scoring process to see which forecast is performing the best. Your `model_id` will be used to track the submissions so any new forecast model requires a new `model_id`.
2. The form gives consent for submissions to be included in Challenge-wide syntheses being carried out by the Challenge organisers. Partipants in the Challenge will be invited to join the synthesis projects on an opt-in basis. 

Questions about Challenge registration and synthesis participation can be directed to [Freya Olsson](mailto:freyao@vt.edu).


# __Decision makers' request__

Lake and reservoir managers could use forecasts of chlorophyll concentrations in surface waters for a number of reasons. For example, a forecast at 1-30 days ahead could be used to:

* To inform swimming/boating restrictions for safe recreation (given the toxin production by some algae species - chlorophyll a can be a useful proxy)
* Prepare water treatment processes - buy/replace filters, buy chemicals, change staffing levels, switch to an alternate drinking water supply

Using the baseline model introduced above, the lake managers are interested in a 1-30 day ahead forecast for July 2023 (and beyond if you like) to understand when a bloom might occur in the next month to inform the open/close status of the lake for swimming and to organise the drinking water treatment processes. A lake will be closed for swimming if concentrations go above 20 mg/L. 

The forecasts should address the following questions:

  + should one lake be chosen for drinking water this month over the other? (how do the forecasts of chlorophyll concentration differ between the sites? which is higher/lower?)
  + what are the maximum concentrations that will occur in the next 30 days? And when will this maximum level occur?
  + are chlorophyll concentrations likely to be higher or lower than normal for the time of year? (comparison with a climatology forecast)
  + what are the chances that the lake(s) will be closed in the next month for swimming?

New data will be available later in the week to assess the success of your forecasts.  

Note: These lakes monitored by NEON are not used for drinking water and the threshold indicated above is for illustrative purposes only!
