---
title: "Sockeye Salmon Returns for Early Stewart Stock 2023" 
format:
  revealjs: 
    theme: moon
slide-level: 3
editor: visual
---

## Group 1

Group members: Jake Lawlor, Yiran Wang, Emmerson Wilson, and Garland Xie

## 1. Basic Ricker Model

::: {.content-visible unless-format="revealjs"}
The Ricker model is a discrete stock-recruitment model that incorporates population growth ($\alpha$) and carrying capacity ($\beta$) for Pacific Salmon. Of all the available stocks, Early Stewart population was randomly chosen since it is one of many that is sensitive to environmental changes (see below models).

It assumes the chosen population experiences density dependence.
:::

Consider the following model:

$$
R_i = \alpha S_i e^{-\beta S_i} 
$$ {#eq-1}

population growth ($\alpha$) and carrying capacity ($\beta$)

---


We can linearize to obtain simpler parameters:

$$
ln(R_i) = ln(\alpha)+ ln(S_i)-S_i/S_{max}
$$ {#eq-2}

---


We place priors on the following:

$$
\widehat{lnR_i} \sim normal(ln(R_i), \tau)
$$ {#eq-3}

$$
ln(\alpha) \sim normal(\mu_{\alpha}, \tau{\alpha}) 
$$ {#eq-4}

$$
ln(S\_{max}) \sim normal(\mu_{Smax}, \tau_{Smax})
$$ {#eq-5}

$$
\tau \sim gamma(\theta, \theta)
$$ {#eq-6}

## 2. Alternate Models {.scrollable}

Power Model

Power Model with environmental covariate

## Alternate 1: Power model {.scrollable}

::: {.content-visible unless-format="revealjs"}
This alternate model is basically shows that log(recruiters) depends on log(spawners). This is also known as a simple power model which does not assume density dependence.
:::

based on the assumption that log(recruiters) depends on log(spawners)

$$
R_i = A S_i^B
$$ {#eq-10}

---

We can linearize equation 10 to the following: $$
ln(R_i) = A+B*ln(S_i)
$$ {#eq-11}

---

This model form has normal priors for both $A$ and $B$:

$$
A \sim normal(\mu_{A}, \tau_{A}) 
$$ {#eq-12} $$
B \sim normal(\mu_{B}, \tau_{B})
$$ {#eq-13}

And the same gamma prior on precision as the above models (equation 6).

::: {.content-visible unless-format="revealjs"}
Note: Jake Lawlor wrote an exploratory R script where he found a significant 1:1 relationship using the lm() function. The explanatory power is 69% (R2). Seems like a promising start for trying to forecast the Early Stewart population!
:::

## Alternate 2: Power model with environmental covariates

This alternate model is an extension to the previous power model that includes the average sea surface temperatures as an environmental covariate:

$$
R_i = A S_i^{B+gEsst_{i}}
$$ {#eq-14}

---

We can linearize this as the following: $$
ln(R_i) = A+B*ln(S_i)+gEsst_{i}
$$ {#eq-15}

--- 

We used the same priors for $A$, $B$, as previous above models and a normal (flat) prior for $g$:

$$
g \sim normal(\mu_g, \tau_g )
$$ {#eq-9}

## 3. Model Comparisons

## Create retroactive predictions

Predict between Years: 2006 - 2021

```{r}
# put graph of models being compared here
```

## Performance Metrics

Metrics Used: Number of times true value fell outside of confidence intervals

what else?

```{r}
# put metric results here
```

## 4. Model Suggestion

Which? Power Model with Environmental Covariate

## 5. Final Forecast

### For 2022 and 2023

```{r}
# put graph with predictions here
```

## Assessment of Stock

Compared to global historical mean

Yellow = prediction within +/-0.5sd of global historical mean

Green = prediction larger than global historical mean by more than +0.5sd

Red = prediction smaller than global historical mean by more than -0.5sd

```{r}
# put graph with yellow green red sections here
```

# LIMIT FISHING TO THE EXTENT POSSIBLE
